---
title: "Introduction to LazymanScript"
author: "Dr. Chia-Hsin Liu"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to LazymanScript}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(LazymanScript)
```

## Introduction

`LazymanScript` provides robust utility functions to seamlessly convert clinical data tables between wide and long formats. Unlike standard pivot functions, `LazymanScript` preserves exact data types (numeric, character, logical) during transformations, making it ideal for messy biomedical datasets. It also includes a robust identity checker to verify that data integrity is maintained during these transformations.

## 1. Converting Wide Data to Long Data

Often, clinical datasets are wide (one row per patient/timepoint). We can convert them to a long format that stores the variable type explicitly.

```{r wide-to-long}
# Create complex sample wide data (150 rows, 25 columns)
set.seed(123)
n_rows <- 150

df <- data.frame(
  s_id = 1:n_rows,
  tp_id = sample(c("Baseline", "Week_12", "Week_24"), n_rows, replace = TRUE),
  age = sample(20:80, n_rows, replace = TRUE),
  gender = sample(c("Male", "Female"), n_rows, replace = TRUE),
  active_smoker = sample(c(TRUE, FALSE, NA), n_rows, replace = TRUE, prob = c(0.2, 0.7, 0.1)),
  height_cm = rnorm(n_rows, mean = 170, sd = 10),
  weight_kg = rnorm(n_rows, mean = 70, sd = 15),
  bmi = numeric(n_rows),
  blood_pressure_systolic = rnorm(n_rows, mean = 120, sd = 15),
  blood_pressure_diastolic = rnorm(n_rows, mean = 80, sd = 10),
  heart_rate = sample(60:100, n_rows, replace = TRUE),
  cholesterol = rnorm(n_rows, mean = 200, sd = 30),
  glucose = rnorm(n_rows, mean = 95, sd = 10),
  wbc = rnorm(n_rows, mean = 7.5, sd = 2.0),
  hemoglobin = rnorm(n_rows, mean = 14, sd = 1.5),
  platelets = sample(150:400, n_rows, replace = TRUE),
  symptom_fatigue = sample(c(TRUE, FALSE), n_rows, replace = TRUE),
  symptom_nausea = sample(c(TRUE, FALSE), n_rows, replace = TRUE),
  treatment_group = sample(c("Control", "Treatment_A", "Treatment_B"), n_rows, replace = TRUE),
  prior_medication = sample(c("Drug_X", "Drug_Y", "None", NA), n_rows, replace = TRUE),
  notes = sample(c("Patient feels well", "Mild headache reported", NA), n_rows, replace = TRUE),
  protocol_deviation = sample(c(TRUE, FALSE), n_rows, replace = TRUE, prob = c(0.05, 0.95)),
  # Columns with all NAs to test edge cases
  dropped_lab_test = rep(NA_real_, n_rows),
  missing_string_data = rep(NA_character_, n_rows),
  unused_flag = rep(NA, n_rows) # Logical NA
)
df$bmi <- df$weight_kg / (df$height_cm / 100)^2

# Introduce some random NAs to numeric columns
df$cholesterol[sample(1:n_rows, 15)] <- NA

# Preview the wide data structure
str(df)

# Convert to long
long_df <- convert_wide_to_long(df, id_cols = c("s_id", "tp_id"))

# View the head of the long data
head(long_df, 15)
```

As you can see, the `convert_wide_to_long` function creates three specific value columns (`value_num`, `value_char`, `value_logic`) and explicitly notes the `value_type` for each feature.

## 2. Restoring Long Data to Wide Data

When you need to pivot back to a wide format for statistical modeling, `LazymanScript` automatically restores the correct R data types based on the `value_type` column!

```{r long-to-wide}
restored_df <- convert_long_to_wide(long_df, id_cols = c("s_id", "tp_id"))

# Reorder columns to match the original dataframe (since dcast sorts them alphabetically)
restored_df <- restored_df[, names(df)]

# View the head of the restored data
head(restored_df, 15)

# Check the restored types
str(restored_df)
```

## 3. Verifying Data Integrity

To ensure that zero data loss or type coercion occurred during your workflow, you can use the `check_data_identity` function.

```{r check-identity}
# Compare the original dataframe with the restored dataframe
result <- check_data_identity(df, restored_df)

# Check if they are identical
print(result$identical)
```

If any columns were dropped, rows were misaligned, or data types were coerced (e.g., numeric to character), the identity checker will flag it as `FAIL` and list the exact issues.

