---
title: "Introduction to LazymanScript"
author: "Dr. Chia-Hsin Liu"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to LazymanScript}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(LazymanScript)
```

## Introduction

`LazymanScript` provides robust utility functions to seamlessly convert clinical data tables between wide and long formats. Unlike standard pivot functions, `LazymanScript` preserves exact data types (numeric, character, logical) during transformations, making it ideal for messy biomedical datasets. It also includes a robust identity checker to verify that data integrity is maintained during these transformations.

## 1. Converting Wide Data to Long Data

Often, clinical datasets are wide (one row per patient/timepoint). We can convert them to a long format that stores the variable type explicitly.

```{r wide-to-long}
# Create sample wide data
df <- data.frame(
  s_id = c(1, 2),
  tp_id = c("A", "B"),
  age = c(25, 30),
  name = c("John", "Jane"),
  active = c(TRUE, FALSE)
)

print(df)

# Convert to long
long_df <- convert_wide_to_long(df, id_cols = c("s_id", "tp_id"))
print(long_df)
```

As you can see, the `convert_wide_to_long` function creates three specific value columns (`value_num`, `value_char`, `value_logic`) and explicitly notes the `value_type` for each feature.

## 2. Restoring Long Data to Wide Data

When you need to pivot back to a wide format for statistical modeling, `LazymanScript` automatically restores the correct R data types based on the `value_type` column!

```{r long-to-wide}
restored_df <- convert_long_to_wide(long_df, id_cols = c("s_id", "tp_id"))

# Reorder columns to match the original dataframe (since dcast sorts them alphabetically)
restored_df <- restored_df[, names(df)]

print(restored_df)

# Check the restored types
str(restored_df)
```

## 3. Verifying Data Integrity

To ensure that zero data loss or type coercion occurred during your workflow, you can use the `check_data_identity` function.

```{r check-identity}
# Compare the original dataframe with the restored dataframe
result <- check_data_identity(df, restored_df)

# Check if they are identical
print(result$identical)
```

If any columns were dropped, rows were misaligned, or data types were coerced (e.g., numeric to character), the identity checker will flag it as `FAIL` and list the exact issues.

